---
title: "3c. Modeling with a Random Forest"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling with a Random Forest}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi       = 100,
  out.width = "95%"
)
```

<hr>

# Summary

-   [Theory](#sec-theory)
-   [Overview](#sec-overview)
-   [Fitting a temporal random forest](#sec-fit)
-   [Threshold selection](#sec-threshold)
-   [Variable importance](#sec-importance)
-   [E-space performance](#sec-espace)
-   [Projecting predictions](#sec-project)
-   [G-space performance](#sec-gspace)
-   [Next steps](#sec-next)

<hr>

# Theory {#sec-theory}

TemporalModelR operates across two complementary dimensions to support temporally explicit modeling:

-   `E-space` - (Environmental Space) represents a location in terms of its environmental conditions, independent of geography. A species' niche in E-space is its tolerance for a given set of environmental variables.
-   `G-space` - (Geographic Space) represents a location in terms of where it physically sits on the landscape. Every point in G-space has a corresponding location in E-space.

Because environmental conditions change over time, the same point in G-space may move through E-space across years, decades, or seasons. Traditional, temporally-static workflows assume both G-space and E-space to be stable in time. TemporalModelR assumes only that the niche (i.e., the species' tolerance in E-space) is stable across the study period, while the E-space coordinates of any G-space location may change over time. Species observations are therefore matched to E-space data correct in both space and time, the niche is modeled in time-independent E-space. This is done in the [Preprocessing temporally explicit data](preprocessing.html) vignette.

By modeling the niche in time-independent E-space using time-independent training data, we can generate a static prediction of a species' environmental tolerances. We can then project these static E-space predictions back onto G-space at explicit time periods, resulting in dynamic, temporally-explicit ENM predictions of species distributions.

<br>

# Overview {#sec-overview}

`build_temporal_rf()` fits one random forest classifier per cross-validation fold as was created during the [Preprocessing temporally explicit data](preprocessing.html) vignette. Each fold's model is trained on all data outside the fold and evaluated on the held-out fold. Predictors are supplied as a character vector via `model_vars` rather than as a formula, since random forests automatically capture nonlinearities and interactions without needing them to be specified. A threshold is selected on the training data and applied to the continuous probability predictions (the per-tree vote fraction for the presence class) to produce binary suitability output for downstream summarisation.

Random forests are an ensemble of decision trees, each grown on a bootstrap sample of the training data and split using a random subset of predictors at each node. The result is a flexible model that captures nonlinear environmental responses and interactions automatically, but sometimes with the the trade-off of interpretability. Random forests do not produce explicit response curves or coefficients, so effects on the response are visualized through variable importance and partial dependence plots. See [Variable importance](#sec-importance).

This vignette runs the seasonal workflow using the bundled `tmr_partition` and `tmr_absences` objects, which are pre-built outputs of the partitioning and pseudoabsence steps produced by the [Preprocessing temporally explicit data](preprocessing.html) vignette using the same call patterns shown there. The dataset itself is described in [About the Example Dataset](dataset.html).This is the same dataset as is used in each modeling vignette [GLM](modeling_glm.html), [GAM](modeling_gam.html), [Random Forest](modeling_rf.html) or [Hypervolume](modeling_hv.html).


The `randomForest` package is a hard dependency of `build_temporal_rf()` and must be installed before running:

```{r, eval=FALSE}
install.packages("randomForest")
```

```{r}
library(TemporalModelR)
library(terra)

data(tmr_partition, package = "TemporalModelR")

data(tmr_absences,  package = "TemporalModelR")
```

<br>

# Fitting a temporal random forest {#sec-fit}

The minimal call needs the partition, the pseudoabsences, a character vector of predictor variable names, and the time columns. We fit a forest with default settings (500 trees, `mtry = floor(sqrt(p))`) and let the function select per-fold thresholds via the True Skill Statistic (TSS), though this threshold may also be set manually or using prevalence. See [Threshold selection](#sec-threshold).

`create_plot = TRUE` allows the function to generate partial dependence plots for each per-fold model result. On the below graphs, variable importance and partial dependence plots are shown per-fold, and the selected threshold is indicated with a horizontal dashed line.

```{r}
rf_out <- build_temporal_rf(
  partition_result     = tmr_partition,
  pseudoabsence_result = tmr_absences,
  model_vars           = c("elevation", "forest_cover", "prseas"),
  rf_params            = list(),
  threshold_method     = "tss",
  output_dir           = file.path(tempdir(), "RF_Models"),
  create_plot          = TRUE,
  time_cols            = c("year", "season"),
  verbose              = FALSE
)
```

The partial dependence plots show how the model captures each predictor's marginal effect on suitability, averaged over the other predictors. The same `create_plot = TRUE` function also visualizes time independent model assessment metrics relevant to the random forest (above). See [E-space performance](#sec-espace).

```{r}
class(rf_out)

names(rf_out)
```

The returned object is a list of class `"TemporalRF"` containing the following objects:

-   `$models` - named list of `randomForest` objects, one per fold (`fold1`, `fold2`, ...). Each is a standard `randomForest::randomForest` fit and accepts the usual `predict()`, `print()`, and `randomForest::importance()` calls.
-   `$thresholds` - named numeric vector of probability thresholds used to binarise predictions for each fold.
-   `$threshold_method` - character or numeric value recording how the thresholds were chosen (`"tss"`, `"prevalence"`, or a fixed numeric value).
-   `$model_vars` - character vector of the predictor variable names supplied to the function. Used by `generate_spatiotemporal_predictions()` to know which raster layers to assemble at projection time.
-   `$variable_importance` - named list of per-fold variable importance tables. Each is a data frame with columns `variable`, `mean_decr_acc` (mean decrease in accuracy from permutation), and `mean_decr_gini` (mean decrease in Gini impurity). See [Variable importance](#sec-importance).
-   `$fold_training_data` - named list of the training data frames used for each fold, with presences and pseudoabsences combined and environmental columns attached.
-   `$fold_test_metrics` - data frame of per-fold E-space metrics computed on the held-out test set. See [E-space performance](#sec-espace).
-   `$output_dir` - path to the directory where the saved RDS of fitted models and the fold test metrics CSV were written.
-   `$model_type` - `"rf"`, used by `generate_spatiotemporal_predictions()` to choose the correct projection logic.
-   `$plots` - list of recorded base-R plots produced when `create_plot = TRUE` (partial dependence plots and the fold metrics bar chart).

Note that unlike GLMs and GAMs, the random forest builder does not take a `model_formula` or `link` argument. Interactions and nonlinearities are captured automatically by the tree structure, and the probability scale comes directly from class vote fractions rather than a link function.

Anything that would normally be passed to `randomForest::randomForest()` can go in `rf_params`. See `?randomForest::randomForest()` for the full list.

<br>

# Threshold selection {#sec-threshold}

Random forests produce probabilities on the 0–1 scale via the vote fraction for the presence class across all trees in the forest. The rest of the TemporalModelR pipeline operates on binary suitability rasters, so a probability threshold must be chosen to convert probabilities into 0/1 predictions. `threshold_method` supports three options:

-   `"tss"` (default) - selects the threshold that maximises True Skill Statistic (sensitivity + specificity − 1) on the training data. See [E-space performance](#sec-espace). This balances commission and omission.
-   `"prevalence"` - sets the threshold equal to the prevalence (proportion of presences) in the training data for that fold.
-   A numeric value between 0 and 1 - used as a fixed threshold for all folds, e.g. `threshold_method = 0.4`.

The chosen thresholds for our four folds:

```{r}
rf_out$thresholds
```

These are per-fold values because each fold trains on a different subset of data and may end up with different probability distributions. Using a single global threshold across folds risks underclassifying suitability in some folds.

<br>

# Variable importance {#sec-importance}

Random forests give per-fold variable importance via the mean decrease in classification accuracy when each predictor is permuted, and via the mean decrease in Gini impurity at the splits each predictor participates in. The function records both in `$variable_importance`:

```{r}
rf_out$variable_importance$fold1
```

Comparing across folds tells you whether the same predictors are consistently informative, or whether their relative ranking changes between spatially or temporally distinct folds. Predictors that flip rank dramatically across folds often indicate that the model is detecting fold-specific patterns rather than a stable niche signal.

When `create_plot = TRUE` was set in [Fitting a temporal random forest](#sec-fit), the function additionally produces partial dependence curves for each predictor, showing the marginal effect of that variable averaged over all others.

<br>

# E-space performance {#sec-espace}

Each fold's held-out test set provides a set of presence and pseudoabsence points the model has never seen based on the folds defined in the [Preprocessing temporally explicit data](preprocessing.html) vignette. We compare predicted vs observed at those points and compute confusion-matrix metrics. Because these are evaluated in E-space (the species' environmental tolerance, independent of geography or time), they give a stable picture of how well the model is preforming.

```{r}
rf_out$fold_test_metrics
```

Columns:

-   `Fold` - fold identifier matching the folds in `tmr_partition$points_sf$fold`.
-   `Threshold` - the per-fold probability threshold used to binarise predictions.
-   `Testing_TP` - count of test-set presence points correctly classified as suitable (true positives).
-   `Testing_FN` - count of test-set presence points incorrectly classified as unsuitable (false negatives).
-   `Testing_TN` - count of test-set pseudoabsence points correctly classified as unsuitable (true negatives).
-   `Testing_FP` - count of test-set pseudoabsence points incorrectly classified as suitable (false positives).
-   `Sensitivity` - proportion of test-set presences correctly classified, `Testing_TP / (Testing_TP + Testing_FN)`. Values close to 1 mean the model rarely misses presences.
-   `Specificity` - proportion of test-set pseudoabsences correctly classified, `Testing_TN / (Testing_TN + Testing_FP)`. Values close to 1 mean the model rarely flags pseudoabsence points as suitable.
-   `TSS` - True Skill Statistic, `Sensitivity + Specificity - 1`. Ranges from -1 to 1, with 0 indicating performance no better than random and values above ~0.4 considered useful for most applications.
-   `Kappa` - Cohen's kappa statistic adjusting overall classification accuracy for the accuracy expected by chance. Ranges from -1 to 1 with similar interpretation to TSS.
-   `AUC` - area under the ROC curve, computed on continuous probabilities rather than binarised predictions, so it is threshold-independent. Useful for comparing the models overall preformance independent of threshold.

The full ROC curve and each above metric are also graphed as an output of `create_plot = T` in the [Fitting a temporal random forest](#sec-fit) section.

E-space metrics are robust to imbalanced sample sizes across time, because they pool across the time series. They are a good metric for assessing overall model fit. Random forest AUC is typically higher than GLM or GAM AUC on the same data, which can be a sign of genuinely better performance or of overfitting. Compare against the G-space metrics computed during projection to distinguish the two. Time-specific (G-space) metrics can also be assessed later when we project the model to spesific G-space and time combinations.

<br>

# Projecting predictions {#sec-project}

`generate_spatiotemporal_predictions()` projects each fold's model onto a stack of environmental rasters matching each requested time step, producing one fold-vote raster per time step. The same call works for all four model types; `model_type` in the model object tells the function which projection logic to use.

For this example we project across all 15 years and 4 seasons, producing 60 prediction layers total, each summarizing votes across the four folds:

```{r}
scaled_dir <- system.file("extdata/rasters_scaled", package = "TemporalModelR")

time_steps <- expand.grid(
  year             = 1:15,
  season           = c("Spring", "Summer", "Autumn", "Winter"),
  stringsAsFactors = FALSE
)

preds <- generate_spatiotemporal_predictions(
  partition_result     = tmr_partition,
  model_result         = rf_out,
  pseudoabsence_result = tmr_absences,
  raster_dir           = scaled_dir,
  variable_patterns    = c(
    "elevation"    = "elevation",
    "forest_cover" = "forest_cover_YEAR",
    "prseas"       = "prseas_YEAR_SEASON"
  ),
  time_cols            = c("year", "season"),
  time_steps           = time_steps,
  output_dir           = file.path(tempdir(), "RF_Predictions"),
  overwrite            = TRUE,
  verbose              = FALSE
)
```

We can now visualize the 60 prediction rasters as a year-by-season grid. Each cell shows the number of folds (0 to 4) that classified that pixel as suitable.

`terra::plot()` limits each call to 16 layers, so we render the stack in four blocks of four years each.

```{r fig.width=10, fig.height=5}
pred_stack <- terra::rast(preds$prediction_files)

pred_names    <- basename(preds$prediction_files)
pred_seasons  <- sub(".*_(Spring|Summer|Autumn|Winter)\\.tif$", "\\1", pred_names)
pred_years    <- as.numeric(sub(".*_(\\d+)_(Spring|Summer|Autumn|Winter)\\.tif$",
                                "\\1", pred_names))
season_levels <- c("Spring", "Summer", "Autumn", "Winter")
stack_order   <- order(pred_years, match(pred_seasons, season_levels))

pred_stack <- pred_stack[[stack_order]]
ordered_years   <- pred_years[stack_order]
ordered_seasons <- pred_seasons[stack_order]
names(pred_stack) <- paste0("Y", ordered_years, "_", ordered_seasons)

block1 <- which(ordered_years %in%  1:4)
block2 <- which(ordered_years %in%  5:8)
block3 <- which(ordered_years %in%  9:12)
block4 <- which(ordered_years %in% 13:15)
```

```{r fig.width=10, fig.height=5}
terra::plot(pred_stack[[block1]], nr = 4, nc = 4,
            mar = c(1.0, 1.0, 1.5, 3.0), legend = FALSE)
```

```{r fig.width=10, fig.height=5}
terra::plot(pred_stack[[block2]], nr = 4, nc = 4,
            mar = c(1.0, 1.0, 1.5, 3.0), legend = FALSE)
```

```{r fig.width=10, fig.height=5}
terra::plot(pred_stack[[block3]], nr = 4, nc = 4,
            mar = c(1.0, 1.0, 1.5, 3.0), legend = FALSE)
```

```{r fig.width=10, fig.height=4}
terra::plot(pred_stack[[block4]], nr = 3, nc = 4,
            mar = c(1.0, 1.0, 1.5, 3.0), legend = FALSE)
```

These rasters represent consensus votes among each of our four folds, with yellow pixels having strong positive consensus among all folds (all four identify the pixel as suitbale) and blue pixels having low consensus among folds (few to no folds identify the given pixel as suitable). We also see that the models correctly visually show one of the main temporal trends in the data: loss of habitat through deforestation starting around year 6.

Visualizing our predictions across each season and year, we see more inconsistant agreement among models than in other model methods presented in the modeling vingettes. RF models have an ability to fit much more complex relationships to variables compared to GLM and GAM, but lack an ability to extrapolate which both of those models have. As a result, RF may be prone to overfitting and this should be watched for. The per-fold partial dependence plots produced when `create_plot = TRUE` are the most direct way to check for this issues, with erratic curves indicating overfitting and curves that don't drop to a clear low value at the edges of the rug indicating extrapolation risk. We see some indication of this above and it manifests in the resulting predictive maps.

The random forest model was unable to find a clean precipitation cutoff, and as a result the predictions vary erratically across folds rather than producing a stable suitability surface. This likewise has to do with the distribution of our absence data.This likewise has to do with the distribution of our absence data, which is restricted to the same Spring and Autumn time steps as the presences and so cannot teach the model what unsuitable precipitation looks like in the other seasons, the same issue seen in our [GLM](modeling_glm.html) and [GLM](modeling_gam.html) models but manifest in a different way. We can explore if other modeling methods are able to fix this hurdle in other vignettes. Alternatively, we could use an alternative method for generating absence points which we describe more in the the [Preprocessing](preprocessing.html) vignette.

Note that here we show that `time_steps` can handle making predictions from compound variables or variables at different temporal scales so long as their temporal scales are nested. For example here "elevation" has no temporal value and is considered to be static across all time steps. "forest_cover" is measured annually, but is considered to be static across all seasons within a year for the purposes of predictions. "prseason" is measured both by year and season, so resulting seasonal predictions reflect that. However if precipitation was only measured based on aggregate seasons but had no associated year, predictios would fail. Predictions can also be made where all variables share the same time step- for example annuual forest cover, annual temprerature, and annual precipitation.

Additionally, a plain vector like `time_steps = 1:15` produces one prediction per value of the first time column; the other time columns are filled in with every unique value present in the occurrence data. So a vector input with `time_cols = c("year", "season")` would produce one prediction per year-season combination observed in the data. A data frame like the `expand.grid()` above gives explicit control: any combination of values can be requested, including ones not present in the occurrence data, as long as the matching rasters exist (Such as Summer or Winter predictions).

<br>

# G-space performance {#sec-gspace}

Once predictions are projected, per-time-step metrics become available. These are computed by overlaying the held-out test points (presences plus, optionally, pseudoabsences) on the prediction raster for each time step they fall in and counting correct vs incorrect classifications.

```{r}
head(preds$timestep_metrics)
```

Columns:

-   `Fold` - fold identifier; one row per fold per time step.
-   `Pct_Suitable` - proportion of the study area predicted suitable in that time step.
-   `N_Pres` - number of held-out presence test points falling in that time step for that fold.
-   `TP` - true positives, presences correctly classified as suitable in that time step.
-   `FN` - false negatives, presences incorrectly classified as unsuitable in that time step.
-   `Sensitivity` - `TP / (TP + FN)` for that fold-time step.
-   `CBP` - cumulative binomial probability. Under the null that each test point lands in suitable area at the rate `Pct_Suitable`, this is the probability of observing exactly `TP` correctly classified points by chance. Small values indicate predictions are better than random.
-   `N_Abs` - number of held-out pseudoabsence test points falling in that time step. `NA` for presence-only models.
-   `TN` - true negatives, pseudoabsences correctly classified as unsuitable.
-   `FP` - false positives, pseudoabsences incorrectly classified as suitable.
-   `Specificity` - `TN / (TN + FP)` for that fold-time step.
-   `TSS` - `Sensitivity + Specificity - 1` for that fold-time step.
-   `year` and `season` - the values of `time_cols` for that row, identifying which time step the metrics belong to.

G-space metrics are powerful in time steps with substantial sample sizes but lose meaning when few records exist (as with out example). A fold with only one presence in a given year can score sensitivity 0 or 1 with no real information content. Likewise, the ability to earn a significant G-space CBP score explicitly depends on sample size. Report E-space (from `$fold_test_metrics`) and G-space (from `$timestep_metrics`) together for a complete view of model performance.

The overall summary gives a per-fold summary across all time steps:

```{r}
preds$overall_summary
```

Rapid visual assessment of these metrics can be done using the function `plot_model_assessment()` which generates simple summary plots for each.

```{r}
plot_model_assessment(
  predictions         = preds,
  time_column         = c("year", "season"),
  secondary_time_mode = "combine",
  model_result        = rf_out
)
```

This produces per-fold time series of percent suitable, sensitivity, specificity, CBP, TP/FN, and TN/FP. When `model_result` is also supplied, overall sensitivity and specificity from `$fold_test_metrics` are added as reference lines. When compound time steps are used (such as year and season), you must choose how they are visualized. Choosing `secondary_time_mode = "combine"` will roll them into one continuous x axis. `secondary_time_mode = "facet"` produces stacked plots as seen below, where the first `time_col` is displayed as the x axis, and a differnt plot is made for each secondary variable (`season` here). By default, the threshold for which CBP is identified as signifcant is 0.05, but this may also be adjusted mannually.

```{r fig.height=20}
plot_model_assessment(
  predictions         = preds,
  time_column         = c("year", "season"),
  secondary_time_mode = "facet",
  model_result        = rf_out,
  cbp_threshold       = 0.001
)
```

<br>

# Next steps {#sec-next}

Now that predictions have been generated, you can assess the model and see if it is preforming satisfactory enough for what your goals are. If this is the case, you can preform postprocessing analyses to try to gain additional inference about the spatiotemporal patterns of change in the study region See the [Postprocessing predictions](postprocessing.html) vignette.

For comparison with other algorithms applied to the same dataset:

-   [Modeling with a GLM](modeling_glm.html) - linear and polynomial responses.
-   [Modeling with a GAM](modeling_gam.html) - smooth nonlinear responses.
-   [Modeling with a Hypervolume](modeling_hv.html) - presence-only n-dimensional kernel density or one-class SVM.
